Phonon polaritonics in two-dimensional materials

### 2019-03-18 paper.tex
edits from reviewers

\newcommand{\HarvardSEAS}{John A. Paulson School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
\newcommand{\MITPhy}{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA}

@@

Before moving on to analyze the electrodynamics of 2D phonon polaritons, we make two comments on lack of LO--TO splitting in 2D polar materials. The first is that this situation is in stark contrast to the situation of polar materials in 3D, which have a finite LO--TO splitting at the $\Gamma$ point. In the absence of such LO--TO splitting in 3D, there would be no frequency compatible with the existence of a phonon  polariton. On the contrary, we will show that in 2D, despite the absence of LO--TO splitting at the $\Gamma$ point, there persists a strongly confined evanescent electromagnetic mode with a high local density of states which in all respects is similar to a phonon polariton of a thin film, but is in fact the 2D LO phonon of the polar monolayer (thus the phrases ``phonon polariton'' and ``2D LO phonon'' may be used somewhat interchangeably as is the case in plasmonics where the terms ``plasmon polariton'' and ``2D plasmon'' are often used interchangeably).

@@

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit being pushed to high wavevectors. This is in contrast with the true monolayer, in which there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. Another anomaly between 2D and 3D is that the dispersion relation of the 2D LO phonon is non-monotonic, and the frequency starts to decrease for sufficiently large wavevector. This can never happen by extrapolating to a monolayer a dielectric function which has finite LO--TO splitting. This turn-back point corresponds to a frequency of around \SI{1500}{\per\cm} and a phonon wavelength of less than \SI{1}{\nm}. Thus, this deviation is challenging to probe in typical optical experiments. However it means that for the monolayer, the range of accessible phonon polaritonic frequencies for optical applications is reduced compared to thick films. Finally, while we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from this limit.

@@

The extreme confinement of electromagnetic fields offered by the 2D phonon polariton presents an opportunity for quantum optical applications in which one seeks to couple an external emitter such as an atom, molecule, defect, or artificial atomic system to electromagnetic fields. Applications of these couplings are ultra-bright single- or two-photon sources, realizing the strong-coupling regime and the associated phenomenology of Rabi oscillations and polaritons, or resolving spectroscopically ``forbidden'' transitions~\cite{koppens2011graphene,autore2018boron,rivera2016shrinking,rivera2017making,kurman2018control} to achieve near-field spectroscopies with momentum and angular momentum resolution not accessible in far-field spectroscopies.

>>>>

\newcommand{\HarvardSEAS}{John A. Paulson School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, 02139, USA}
\newcommand{\MITPhy}{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, 02139, USA}

@@

Before moving on to analyze the electrodynamics of 2D phonon polaritons, we make three comments on lack of LO--TO splitting in 2D polar materials. The first is that this situation is in stark contrast to the situation of polar materials in 3D, which have a finite LO--TO splitting at the $\Gamma$ point. In the absence of such LO--TO splitting in 3D, there would be no frequency compatible with the existence of a phonon  polariton. On the contrary, we will show that in 2D, despite the absence of LO--TO splitting at the $\Gamma$ point, there persists a strongly confined evanescent electromagnetic mode with a high local density of states which in all respects is similar to a phonon polariton of a thin film, but is in fact the 2D LO phonon of the polar monolayer (thus the phrases ``phonon polariton'' and ``2D LO phonon'' may be used somewhat interchangeably as is the case in plasmonics where the terms ``plasmon polariton'' and ``2D plasmon'' are often used interchangeably).

@@

The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~ \cite{sohier2017breakdown}, one may describe the LO-TO splitting in terms of Eq. ~\ref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO-TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $\frac{e^2K_0(qr)}{2\pi\epsilon_0}$, where $K_0$ is a modified Bessel function of the second kind and of order zero. Thus, the $q$-dependence of the LO-TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\ref{eq:2dsusceptibility}.

%\textcolor{blue}{The third comment is that the considerations of this section can be extended to few-layer systems, such as hBN bilayers, which may still be considered 2D for polariton wavelengths large compared to the bilayer thickness. The considerations may also be extended to 1D polar materials, such as BN nanotubes. For bilayers, as was discussed in Ref.~ \cite{sohier2017breakdown}, one may describe the LO-TO splitting in terms of Eq. ~\ref{eq:zerosofresponse}, with the 2D Coulomb interaction, but with eigendisplacements appropriate to the few layer system. For a small number of layers $N$, this should enhance the LO-TO splitting by a factor of $N$. For 1D systems, beyond using the eigendisplacements appropriate to the 1D system, one must also use the 1D Coulomb interaction, cutoff by a length $R$ which for a nanotube, would be its radius. In that case, the 1D Coulomb interaction is $\frac{e^2K_0(qr)}{2\pi\epsilon_0}$, where $K_0$ is a modified Bessel function of the second kind and of order zero. Thus, the $q$-dependence of the LO-TO splitting is $q^2K_0(qR)$. This is in close analogy with the relation for the plasma frequency of 1D plasmons, as expected from the considerations of the previous paragraph. As regards the optical properties, the optical conductivity is still governed by Eq.~\ref{eq:2dsusceptibility}.}

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit being pushed to high wavevectors. This is in contrast with the true monolayer, in which there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. The bulk dielectric function of Eq.~\ref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of 12 nm, which is very short, the polariton frequency is only a few cm$^{-1}$ above the TO frequency, within the damping linewidth of the TO frequency. %\textcolor{blue}{The bulk dielectric function of Eq.~\ref{eq:3deps} fails to capture this lack of higher-order modes, and thus hyperbolicity. For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Fig.~\ref{fig:2}(c), because even for a polariton wavelength of 12 nm, which is very short, the polariton frequency is only a few cm$^{-1}$ above the TO frequency, within the damping linewidth of the TO frequency.}

Another anomaly between 2D and 3D is that the dispersion relation of the 2D LO phonon is non-monotonic, and the frequency starts to decrease for sufficiently large wavevector. This can never happen by extrapolating to a monolayer a dielectric function which has finite LO--TO splitting. This turn-back point corresponds to a frequency of around \SI{1500}{\per\cm} and a phonon wavelength of less than \SI{1}{\nm}. Thus, this deviation is challenging to probe in typical optical experiments. However it means that for the monolayer, the range of accessible phonon polaritonic frequencies for optical applications is reduced compared to thick films. Finally, while we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from this limit. This qualitative similarity is to be somewhat expected, as in many cases, a 2D layer can be considered as an very thin film of bulk, provided that the microscopic properties of the bulk and monolayer are the same. In fact, if we take an thin film of thickness $t$ of hBN of dielectric function given by Eq.~\ref{eq:3deps}, and focus only on the in-plane response, then there is a mathematical condition for the thin film to exactly mirror the bulk. It is: 
\begin{equation}
(\omega^{2 (\mathrm{bulk})}_{\mathrm{LO}} - \omega^{2 (\mathrm{bulk})}_{\mathrm{TO}})t = 4\omega^{(\mathrm{mono})}_{\mathrm{TO}}v^{(\mathrm{mono})}_g,
\end{equation}
where "bulk" denotes a property of the bulk dielectric function and "mono" denotes that of the monolayer conductivity.

%\textcolor{blue}{This qualitative similarity is to be somewhat expected, as in many cases, a 2D layer can be considered as an very thin film of bulk, provided that the microscopic properties of the bulk and monolayer are the same. In fact, if we take an thin film of thickness $t$ of hBN of dielectric function given by Eq.~\ref{eq:3deps}, and focus only on the in-plane response, then there is a mathematical condition for the thin film to exactly mirror the bulk. It is: 
%\begin{equation}
%(\omega^{2 (\mathrm{bulk})}_{\mathrm{LO}} - \omega^{2 (\mathrm{bulk})}_{\mathrm{TO}})t = 4\omega^{(\mathrm{mono})}_{\mathrm{TO}}v^{(\mathrm{mono})}_g,
%\end{equation}
%where "bulk" denotes a property of the bulk dielectric function and "mono" denotes that of the monolayer conductivity.}

@@

The extreme confinement of electromagnetic fields offered by the 2D phonon polariton presents an opportunity for quantum optical applications in which one seeks to couple an external emitter such as an atom, molecule, defect, or artificial atomic system to electromagnetic fields. Applications of these couplings are ultra-bright single- or two-photon sources, realizing the strong-coupling regime and the associated phenomenology of Rabi oscillations and polaritons, or resolving spectroscopically ``forbidden'' transitions~\cite{koppens2011graphene,autore2018boron,rivera2016shrinking,rivera2017making,kurman2018control} to achieve near-field spectroscopies with momentum and angular momentum resolution not accessible in far-field spectroscopies. One expects that a monolayer offers the largest spontaneous emission enhancement compared to any multilayer system for a sufficiently nearby emitter. This assertion is shown in the Supplementary Materials.

%\textcolor{blue}{One expects that a monolayer offers the largest spontaneous emission enhancement compared to any multilayer system for a sufficiently nearby emitter. This assertion is shown in the Supplementary Materials.}
